Galois Slicing as Differentiation

### 2025-07-11 main/approx-as-tangents.tex
Drop indentation in  environment at end of 2.2.
Some leftmargin parameterisation.
[ tex ] further space saving

We motivate our approach to \GPS by showing how to combine ideas from differential geometry and stable domain theory to reconstruct \GPS in a denotational setting. From this basis, we apply the CHAD framework of V{\'a}k{\'a}r \etal to obtain a denotational model of automatic differentiation that gives us an effective way to interpret programs as functions along with their forward and backwards approximation maps.

@@

Smooth functions $f$ between manifolds $M$ and $N$ are functions on their points that are locally differentiable on $\RR^n$. Manifolds and smooth functions form a category $\Man$. Each smooth function induces maps of the (co)tangent spaces:
\begin{itemize}
\item The \emph{forward derivative} (tangent map, pushforward) $\pushf{f}_x$ is a linear map $\tangents_x(M) \linearto \tangents_{f(x)}(N)$. In the Euclidean case when $M = \RR^m$ and $N = \RR^n$, the tangent map can be represented by the Jacobian matrix of partial derivatives of $f$ at $x$.
\item The \emph{backward derivative} (cotangent map, pullback) $\pullf{f}_x$ is a linear map $\cotangents_{f(x)}(N) \linearto \cotangents_x(M)$. In the Euclidean case, the backward derivative is represented by the transpose of the Jacobian of $f$ at $x$.
\end{itemize}

@@

\begin{example}\label{ex:stable-functions}
  \leavevmode\\[-3ex]
  \begin{enumerate}
  \item The function $\mathrm{strictOr}$ is stable. For example, for
    the input-output pair
    $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input
    that gives this output is exactly $(\mathsf{tt}, \mathsf{ff})$. If
    we take the approximation $\bot \leq \mathsf{tt}$ of the output,
    then the corresponding minimal input is $(\bot, \bot)$. The
    function $\mathrm{shortCircuitOR}$ is also stable. For the
    input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$,
    the minimal input that gives this input is $(\mathsf{tt},\bot)$,
    indicating that the presence of $\mathsf{ff}$ in the second
    argument was not necessary to produce this output. As with
    $\mathrm{strictOr}$, the minimal input required to produce the
    output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

@@

\begin{enumerate}
\item Even without completeness, in bounded meet posets and L-posets
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. We discuss this further in
  \secref{conclusion}.) In \exref{intervals-and-maxima-elements},
  maximal elements could be taken to be the ``proper values''. One
  idea is to restrict to conditionally multiplicative or stable
  functions that preserve maximal elements. However, this idea fails
  at higher order: functions that take maximal elements to maximal
  elements are not themselves maximal elements. We could devise a
  category of L-posets with totality predicates (which would pick out
  maximal elements at first-order) and totality preserving functions,
  but we prefer a more direct method of separating values proper from
  their approximations using the \emph{Category of Families}
  construction as we explain in \secref{models-of-total-gps:fam}.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infima of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} show how
  to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

>>>>

We motivate our approach by showing how to combine ideas from differential geometry and stable domain theory to reconstruct the ideas of \GPS in a denotational setting.
% From this basis, we apply the CHAD framework of V{\'a}k{\'a}r \etal to obtain a denotational model of automatic differentiation that gives us an effective way to interpret programs as functions along with their forward and backwards approximation maps.

@@

Smooth functions $f$ between manifolds $M$ and $N$ are functions on their points that are locally differentiable on $\RR^n$. Manifolds and smooth functions form a category $\Man$. Each smooth function induces maps of the (co)tangent spaces:
\begin{itemize}[leftmargin=\enummargin]
\item The \emph{forward derivative} (tangent map, pushforward) $\pushf{f}_x$ is a linear map $\tangents_x(M) \linearto \tangents_{f(x)}(N)$. In the Euclidean case when $M = \RR^m$ and $N = \RR^n$, the tangent map can be represented by the Jacobian matrix of partial derivatives of $f$ at $x$.
\item The \emph{backward derivative} (cotangent map, pullback) $\pullf{f}_x$ is a linear map $\cotangents_{f(x)}(N) \linearto \cotangents_x(M)$. In the Euclidean case, the backward derivative is represented by the transpose of the Jacobian of $f$ at $x$.
\end{itemize}

@@

\begin{example}\label{ex:stable-functions}
  \leavevmode\\[-3ex]
  \begin{enumerate}[leftmargin=\enummargin]
  \item The function $\mathrm{strictOr}$ is stable. For example, for
    the input-output pair
    $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input
    that gives this output is exactly $(\mathsf{tt}, \mathsf{ff})$. If
    we take the approximation $\bot \leq \mathsf{tt}$ of the output,
    then the corresponding minimal input is $(\bot, \bot)$. The
    function $\mathrm{shortCircuitOR}$ is also stable. For the
    input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$,
    the minimal input that gives this input is $(\mathsf{tt},\bot)$,
    indicating that the presence of $\mathsf{ff}$ in the second
    argument was not necessary to produce this output. As with
    $\mathrm{strictOr}$, the minimal input required to produce the
    output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

@@

\begin{enumerate}[leftmargin=*]
\item Even without completeness, in bounded meet posets and L-posets
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. We discuss this further in
  \secref{conclusion}.) In \exref{intervals-and-maxima-elements},
  maximal elements could be taken to be the ``proper values''. One
  idea is to restrict to conditionally multiplicative or stable
  functions that preserve maximal elements. However, this idea fails
  at higher order: functions that take maximal elements to maximal
  elements are not themselves maximal elements. We could devise a
  category of L-posets with totality predicates (which would pick out
  maximal elements at first-order) and totality preserving functions,
  but we prefer a more direct method of separating values proper from
  their approximations using the \emph{Category of Families}
  construction as we explain in \secref{models-of-total-gps:fam}.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infima of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} show how
  to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

### 2025-07-11 main/conclusion.tex
[ tex ] further space saving

\paragraph{Categorical Models of Differentation} We have already identified \conref{tangent-stable-fns} and
\conref{tangent-fam} as potential connections to categorical models of differentation. It would also be
interesting to discover any possible links to Differential Linear Logic
\cite{ehrhard_differential_2006}. Differential Linear Logic and the Dialectica translation have been used to
model reverse differentiation by \citet{kerjean-pedrot2024} that is similar to the CHAD approach
of \citet{vakar22}.

@@

\paragraph{Recursion and Partiality} This work has only examined the case for total programs. Even though we
took stable domain theory as our starting point, we did not make any use of directed completeness or similar
properties of domains. We expect that an account of recursion in an extension of the framework discussed so
far would likely involve families of bounded lattices indexed by DCPOs, where the order of the DCPO would be
reflected in embedding-projection relationships between the lattices. There are also other refinements of
stable domain theory such as \citet{berry79}'s bidomains and \citet{laird07} bistable biorders that separate
the extensional and stable orders on the same set, in a way that might be similar to
\exref{intervals-and-maxima-elements}.

>>>>

% \paragraph{Categorical Models of Differentation} We have already identified \conref{tangent-stable-fns} and
% \conref{tangent-fam} as potential connections to categorical models of differentation.

@@

\paragraph{Recursion and Partiality} This work has only examined the case for total programs. Even though we
took stable domain theory as our starting point, we did not make any use of directed completeness or similar
properties of domains. We expect that an account of recursion in an extension of the framework discussed so
far would likely involve families of bounded lattices indexed by DCPOs, where the order of the DCPO would be
reflected in embedding-projection relationships between the lattices. \citet{berry79}'s bidomains and
\citet{laird07}'s bistable biorders have separate extensional and stable orders on the same set, in a way that
might be similar to \exref{intervals-and-maxima-elements}.

### 2025-07-11 main/related-work.tex
[ tex ] further space saving

\subsection{Stable Domain Theory}

@@

\subsection{Automatic Differentiation}

@@

\subsection{\GPS}
\label{sec:related-work:galois-slicing}

@@

\subsection{Tangent Categories}

\emph{Tangent categories}, due originally to \citet{rosický84} and developed by \citet{cockett14,cockett18},
provide an abstract categorical framework for reasoning about differentiation, inspired by the structure of
the tangent bundle in differential geometry. In a tangent category, each object $X$ is equipped with a tangent
bundle $T(X)$, and each morphism $f: X \to Y$ has a corresponding differential map $T(f): T(X) \to T(Y)$
satisfying axioms analogous to the chain rule and linearity of differentiation. Tangent categories generalise
Cartesian differential categories \cite{cdcs}, which model differentiation over cartesian closed categories
using a syntactic derivative operator. Reverse Tangent categories \cite{reverse-tangents} further axiomatise
the existence of reverse derivatives. In \conref{tangent-stable-fns} and \conref{tangent-fam}, we have
conjectured that the categories we have identified in this paper as models of \GPS are Tangent categories. As
well as providing some substantiation to the claim that \GPS is a form of differentiation, this would also
clarify the role of higher derivatives in \GPS, which we conjecture are related to \emph{program
  differencing}.

% \subsection{Differential Linear Logic}

% Differential Linear Logic \cite{dill} is an extension of Linear Logic with a derivative operator that acts
% like

% \subsection{Lens Categories}

>>>>

\paragraph{Stable Domain Theory}

@@

\paragraph{Automatic Differentiation}

@@

\paragraph{\GPS}
\label{sec:related-work:galois-slicing}

@@

\paragraph{Tangent Categories and Differential Linear Logic}

\emph{Tangent categories}, due originally to \citet{rosický84} and developed by \citet{cockett14,cockett18},
provide an abstract categorical framework for reasoning about differentiation, inspired by the structure of
the tangent bundle in differential geometry. In a tangent category, each object $X$ is equipped with a tangent
bundle $T(X)$, and each morphism $f: X \to Y$ has a corresponding differential map $T(f): T(X) \to T(Y)$
satisfying axioms analogous to the chain rule and linearity of differentiation. Tangent categories generalise
Cartesian differential categories \cite{cdcs}, which model differentiation over cartesian closed categories
using a syntactic derivative operator. Reverse Tangent categories \cite{reverse-tangents} further axiomatise
the existence of reverse derivatives. In \conref{tangent-stable-fns} and \conref{tangent-fam}, we have
conjectured that the categories we have identified in this paper as models of \GPS are Tangent
categories. This would clarify the role of higher derivatives in \GPS, which we conjecture are related to
\emph{program differencing}. There are likely links to Differential Linear Logic
\cite{ehrhard_differential_2006}. Differential Linear Logic and the Dialectica translation have been used to
model reverse differentiation by \citet{kerjean-pedrot2024}.

% \paragraph{Lens Categories}
